NumLin: Linear Types for Linear Algebra

### 2019-05-26 write-up/paper/discussion.tex
Make stuff consistent
Syntax & spacing, with a few small edits from Reviewer C's review.
Incorporate final review feedback
Also reduce warnings.

The original implementation used 6 temporaries, one of which was immediately
spotted as never being used due to linearity. It also contained two variables
which were marked as \texttt{intent(in)} but would have been written over by
calls to `gemm', spotted by the fractional capabilities feature. Furthermore,
it used a matrix \emph{twice} in a call to `symm', once with a read permission
but once with a \emph{write} permission.  Fortran assumes that any parameter
being written to is not aliased and so this call was not only incorrect, but
illegal according to the standard, both aspects of which were captured by
linearity and fractional capabilities.

@@

@@

@@

This approach is similar to the one taken in linear algebra libraries for Rust
-- these libraries typically take advantage of the distinction that Rust's type
system offers between mutable views/references to arrays.  The work of
\cite{weiss} and \cite{rustbelt} suggest that Rust's borrow-checker \emph{can
be expressed in simpler terms} using fractional permissions, though to our
knowledge the programmer-visible lifetime analysis in Rust has never been
formalized.

@@

@@

@@

However, since our type system shows demonstrates unification is ``all one
needs'' for fractions, it should be possible to \emph{encode} \lang's approach
to fractional permissions in Linear Haskell by adding a GADT-style natural
number index to array types tracking the fraction, which should enable
supporting high-performance BLAS bindings in Linear Haskell. Actually
implementing this is something we leave for future work, as there remains one
issue which we do not see a good encoding for. Namely, only having support for
linear functions makes it a bit inconvenient to manipulate linear values
directly -- programs end up taking on a CPS-like structure. This seems to
remain an advantage of a direct implementation of linear types over the Linear
Haskell style.

@@

Going forwards, one may wish to eliminate even more runtime errors from \lang,
by extending its type system. For example, we could have used existential types
to statically track pointer identities~\cite{morrisett}, or parametric
polymorphism.

>>>>

The original implementation used 6 temporaries, one of which was immediately
spotted as never being used due to linearity. It also contained two variables
which were marked as \texttt{intent(in)} but would have been written over by
calls to `gemm', spotted by the fractional permissions feature. Furthermore,
it used a matrix \emph{twice} in a call to `symm', once with a read permission
but once with a \emph{write} permission.  Fortran assumes that any parameter
being written to is not aliased and so this call was not only incorrect, but
illegal according to the standard, both aspects of which were captured by
linearity and fractional permissions.

@@

\subsubsection{Linear types for \emph{implementing} linear algebra routines}

@@

\subsubsection{Our contribution: linear types for enforcing correct
    \emph{usage} of linear algebra routines}

@@

\subsubsection{Traditionally complex approaches to sharing}

Our approach is similar to the one taken in linear algebra libraries for Rust
-- these libraries typically take advantage of the distinction that Rust's type
system offers between mutable views/references to arrays.  The work of
\cite{weiss} and \cite{rustbelt} suggest that Rust's borrow-checker \emph{can
be expressed in simpler terms} using fractional permissions, though to our
knowledge the programmer-visible lifetime analysis in Rust has never been
formalized.

@@

\subsubsection{Our contribution: a simpler approach to sharing}

@@

\subsubsection{Implications}

@@

However, since our type system shows unification is ``all one
needs'' for fractions, it should be possible to \emph{encode} \lang's approach
to fractional permissions in Linear Haskell by adding a GADT-style natural
number index to array types tracking the fraction, which should enable
supporting high-performance BLAS bindings in Linear Haskell. Actually
implementing this is something we leave for future work, as there remains one
issue which we do not see a good encoding for. Namely, only having support for
linear functions makes it a bit inconvenient to manipulate linear values
directly -- programs end up taking on a CPS-like structure. This seems to
remain an advantage of a direct implementation of linear types over the Linear
Haskell style.

@@

Going forwards, one may wish to eliminate even more runtime errors from \lang,
by extending its type system. For example, we could have used existential types
to statically track pointer identities~\cite{ahmed20073}, or parametric
polymorphism.

### 2019-05-26 write-up/paper/formal_system.tex
Fix reference
Incorporate final review feedback
Also reduce warnings.

Using this, we can explain how the ! annotation on variables -- first
introduced in the factorial example in \ref{subsubsec:factorial} -- works. That
is, we can explain why the meaning of $\mathbf{let}\ !x = e\ \mathbf{in}\ e'$
can be expressed using only the rules presented thus far, as $\mathbf{let\
Many}\ x = e\ \mathbf{in\ } \mathbf{let\ Many}\ x = \mathbf{Many}\
(\mathbf{Many}\ x)\ \mathbf{in}\ e'$.\footnote{\emph{Why} we have this at all
    is for the sake of ergonomics when using binary arithmetic operations
    (e.g.\ of type $!\ottkw{int} \, \otimes \, !\ottkw{int} \multimap \;
    !\ottkw{int}$): a programmer should be able to write \highl{let x = 5 + 5
    in x - x}, which, although non-linear in \highl{x}, is morally right
    because integers and operations on them rarely need to be linear. Though it
    should be possible to handle this using a LNL-style presentation of linear
    types~\cite{benton1995} (using adjoint modalities to distinguish between
    intrinsically linear and intrinsically intuitionistic types) that is a
    pretty big digression from the stated goals of this paper.}
The reader is invited to quickly convince themselves that the following
meta-rule is provable using \textsc{Ty\_Bang\_Intro} (twice),
\textsc{Ty\_Bang\_Elim} (twice) and weakening the intuitionistic environment
$\Delta$ (once).

@@

\subsection{Logical Relation}

@@

We then define a step-indexed logical relation in the style of Morrisett et.
al~\cite{ahmed20073}. $(\varsigma, v) \in \V{k}{t}$ means it takes a heap with
exactly $\varsigma$ resources to produce a value $v$ of type $t$ in at most $k$
steps. So, something like a $\ottkw{unit}$ or a $!t$ need no resources, whereas
a $f\, \ottkw{mat}$ needs exactly $f$ ownership of a some matrix and a pair
needs a $\star$ combination of the heaps required for each component.
\begin{align*}
  \V{k}{ \Unit } &= \{ (\empH, \ast) \} \\
  \V{k}{ f \, \Mat } &= \{ (\{ l \mapsto_{2^{-f}} \_ \} , l) \} \\
  \V{k}{ \Bang t } &= \{ (\empH, \Many\, v) \mid (\empH, v) \in \V{k}{t} \} \\
  \V{k}{ t_1 \otimes t_2 } &= \{ (\varsigma_1 \star \varsigma_2, ( v_1, v_2 )) \mid (\varsigma_1, v_1) \in \V{k}{t_1} \wedge (\varsigma_2, v_2) \in \V{k}{t_2} \}
\end{align*}

The definition of $\V{k}{ '\!f\!c.\ t}$ says a value and heap
must be the same regardless of what fraction is substituted into both; the
$k-1$ is to take into account fraction specialisation takes ones step
(\textsc{Op\_Spc}).

@@

To understand the definition of $\V{k}{t' \multimap t}$, we must first look at
$\C{k}{t}$, the computational interpretation of types. Intuitively, it is a
combination of a frame rule on heaps (no interference), type-preservation and
termination (in $j < k$ steps) to either an error or a heap-and-expression,
with the further condition that if the expression is a syntactic value then it
is also one semantically.
\begin{align*}
    \C{k}{ t } &= \{ (\varsigma_s, e_s) \mid \forall \, j < k, \sigma_r.\ \varsigma_s \star \varsigma_r \textrm{ defined } \Rightarrow \langle \sigma_s + \sigma_r, e_s \rangle \rightarrow^j \ottkw{err}\ \vee \exists \sigma_f, e_f.\\
               & \qquad \qquad \langle \sigma_s + \sigma_r, e_s \rangle \rightarrow^j \langle \sigma_f + \sigma_r, e_f \rangle \wedge ( e_f \textrm { is a value } \Rightarrow ( \varsigma_f \star \varsigma_r, e_f ) \in \V{k-j}{t} ) \}
\end{align*}

@@

If an expression is syntactically type-checked, then for an arbitrary number of
steps $k$, under any substitution $\theta$ of free fractional-permission
variables, linear variables $\gamma$ (with a suitable heap $\varsigma$) and
intuitionistic variables $\delta$, that suitable heap and expression
$(\varsigma, \theta(\delta(\gamma(e))))$ are in the computational
interpretation of the type of that expression $\C{k}{ \theta(t) }$.

The \emph{computational interpretation} is as defined before: a combination of
a frame rule on heaps (no interference), type-preservation and termination (in
$j < k$ steps) to either an error or a heap-and-expression, with the further
condition that if the expression is a syntactic value then it is also one
semantically.

It identifies executions that do no un- or ill-defined behaviours (e.g.\ adding
a boolean and an integer). Since our operational semantics explicitly models
deallocation, we now know no well-typed program will ever try to access
deallocated memory, establishing the correctness of our memory management
checking.

@@

The case for \textsc{Ty\_Gen} follows a similar pattern, but has the extra
complication of reducing an expression with an arbitrary fractional permission
variable in it, and then instantiating it at the last momemnt to conclude,
which is where \ref{fracPermSub} (heap-and-expressions take the same steps of
evaluation under any substitution of their free fractional permissions) is
used.

>>>>

Using this, we can explain how the ! annotation on variables -- first
introduced in the factorial example in \ref{subsubsec:factorial} -- works. That
is, we can explain why the meaning of $\mathbf{let}\ !x = e\ \mathbf{in}\ e'$
can be expressed using only the rules presented thus far, as $\mathbf{let\
Many}\ x = e\ \mathbf{in\ } \mathbf{let\ Many}\ x = \mathbf{Many}\
(\mathbf{Many}\ x)\ \mathbf{in}\ e'$.\footnote{\emph{Why} we have this at all
    is for the sake of ergonomics when using binary arithmetic operations
    (e.g.\ of type $!\ottkw{int} \multimap \; !\ottkw{int} \multimap \;
    !\ottkw{int}$): a programmer should be able to write \highl{let x = 5 + 5
    in x - x}, which, although non-linear in \highl{x}, is morally right
    because integers and operations on them rarely need to be linear. Though it
    should be possible to handle this using a LNL-style presentation of linear
    types~\cite{benton1995} (using adjoint modalities to distinguish between
    intrinsically linear and intrinsically intuitionistic types) that is a
    pretty big digression from the stated goals of this paper.}
The reader is invited to quickly convince themselves that the following
meta-rule is provable using \textsc{Ty\_Bang\_Intro} (twice),
\textsc{Ty\_Bang\_Elim} (twice) and weakening the intuitionistic environment
$\Delta$ (once).

@@

\subsection{Logical Relation}\label{subsec:logical_rel}

@@

We then define a step-indexed logical relation in the style of Ahmed et.
al~\cite{ahmed20073}. $(\varsigma, v) \in \V{k}{t}$ means it takes a heap with
exactly $\varsigma$ resources to produce a value $v$ of type $t$ in at most $k$
steps. So, something like a $\ottkw{unit}$ or a $!t$ need no resources, whereas
a $f\, \ottkw{mat}$ needs exactly $f$ ownership of a some matrix and a pair
needs a $\star$ combination of the heaps required for each component.
\begin{align*}
  \V{k}{ \Unit } &= \{ (\empH, \ast) \} \\
  \V{k}{ f \, \Mat } &= \{ (\{ l \mapsto_{2^{-f}} \_ \} , l) \} \\
  \V{k}{ \Bang t } &= \{ (\empH, \Many\, v) \mid (\empH, v) \in \V{k}{t} \} \\
  \V{k}{ t_1 \otimes t_2 } &= \{ (\varsigma_1 \star \varsigma_2, ( v_1, v_2 )) \mid (\varsigma_1, v_1) \in \V{k}{t_1} \wedge (\varsigma_2, v_2) \in \V{k}{t_2} \}
\end{align*}

The definition of $\V{k}{ '\!f\!c.\ t}$ says a value and heap
must be the same regardless of what fraction is substituted into both; the
$k-1$ is to take into account fraction specialisation takes one step
(\textsc{Op\_Spc}).

@@

To understand the definition of $\V{k}{t' \multimap t}$, we must first look at
$\C{k}{t}$, the computational interpretation of types. Intuitively, it is a
combination of a frame rule on heaps (no interference), type-preservation and
termination (in $j < k$ steps) to either an error or a heap-and-expression.
For the case of termination to a heap-and-expression, there is a further
condition: if the expression is a value syntactically then it is also one
semantically.
\begin{align*}
    \C{k}{ t } &= \{ (\varsigma_s, e_s) \mid \forall \, j < k, \sigma_r.\ \varsigma_s \star \varsigma_r \textrm{ defined } \Rightarrow \langle \sigma_s + \sigma_r, e_s \rangle \rightarrow^j \ottkw{err}\ \vee \exists \sigma_f, e_f.\\
               & \qquad \qquad \langle \sigma_s + \sigma_r, e_s \rangle \rightarrow^j \langle \sigma_f + \sigma_r, e_f \rangle \wedge ( e_f \textrm { is a value } \Rightarrow ( \varsigma_f \star \varsigma_r, e_f ) \in \V{k-j}{t} ) \}
\end{align*}

@@

If an expression $e$ is syntactically type-checked (against a type $t$), then
\begin{itemize}
    \item for an arbitrary number of steps $k$,
    \item under any substitution of
        \begin{itemize}
            \item free fractional permission variables $\theta$,
            \item linear variables with a suitable heap $(\gamma, \varsigma)$ and
            \item intuitionistic variables $\delta$,
        \end{itemize}
    \item the aforementioned suitable heap and expression $(\varsigma, \theta(\delta(\gamma(e))))$
    \item are in the computational interpretation $\C{k}{ \theta(t) }$ of the type $t$.
\end{itemize}
The \emph{computational interpretation} is as defined before
(Section~\ref{subsec:logical_rel}); it identifies executions that do no un- or
ill-defined behaviours (e.g.\ adding a boolean and an integer). Since our
operational semantics explicitly models deallocation, we now know no well-typed
program will ever try to access deallocated memory, establishing the
correctness of our memory management checking.

@@

The case for \textsc{Ty\_Gen} follows a similar pattern, but has the extra
complication of reducing an expression with an arbitrary fractional permission
variable in it, and then instantiating it at the last moment to conclude,
which is where \ref{fracPermSub} (heap-and-expressions take the same steps of
evaluation under any substitution of their free fractional permissions) is
used.

### 2019-05-26 write-up/paper/intro.tex
Incorporate final review feedback
Also reduce warnings.

So in practice, programmers face a tradeoff: they can eschew safety
and exploit the full power of the underlying linear algebra libraries,
or they can obtain safety at the price of unneeded copies and worse
memory efficiency. In this work, we show that this tradeoff is not a
fundamental one.

@@

In this paper
\begin{itemize}
    \item we describe \lang, a linearly typed language for linear algebra programs
    \item we illustrate that \lang's design and features are well-suited to its
        intended domain with progressively sophisticated examples
    \item we prove \lang's soundness, using a step-indexed logical relation
    \item we describe a very simple, unification based type-inference algorithm
        for polymorphic fractional permissions (similar to ones used for
        parametric polymorphism), demonstrating an alternative approach to
        dataflow analysis \cite{bierhoff}
    \item we describe an implementation that is both compatible with and usable
        from existing code
    \item we show an example of how using \lang\ helped highlight linearity
        and aliasing bugs, and reduce the memory usage of a \emph{generated}
        linear algebra program
    \item we show that using \lang, we can achieve parity with C for linear
        algebra routines, whilst having much better static guarantees about the
        linearity and aliasing behaviour of our programs.
\end{itemize}

>>>>

So in practice, programmers face a trade-off: they can eschew safety
and exploit the full power of the underlying linear algebra libraries,
or they can obtain safety at the price of unneeded copies and worse
memory efficiency. In this work, we show that this trade-off is not a
fundamental one.

@@

Our contribution is the idea applying of linear types with fractional
permissions to enforce the correct \emph{usage} (as opposed to
\emph{implementation}) of linear algebra libraries. We explain the idea in
detail and provide evidence for its efficacy. Prior type systems for fractional
permissions \cite{boyland2003, bierhoff, bierhoff2008plural} are quite complex.
This is because these type systems typically encode a sophisticated analysis to
automatically infer how fractional permissions should be split and rejoined.

In contrast, in \lang, we made sharing and merging explicit. As a result, we
were able to drastically simplify the type system. Thefore, our formal system
is very close to standard presentations of linear logic, and the implementation
complexity is no worse than that for parametric polymorphism.

In this paper
\begin{itemize}
    \item we describe \lang, a linearly typed language for linear algebra programs
    \item we illustrate that \lang's design and features are well-suited to its
        intended domain with progressively sophisticated examples
    \item we prove \lang's soundness, using a step-indexed logical relation
    \item we describe a very simple, unification based type-inference algorithm
        for polymorphic fractional permissions (similar to ones used for
        parametric polymorphism), demonstrating an alternative approach to
        dataflow analysis \cite{bierhoff}
    \item we describe an implementation that is compatible with and usable
        from existing code
    \item we show an example of how using \lang\ helped highlight linearity
        and aliasing bugs, and reduce the memory usage of a \emph{generated}
        linear algebra program
    \item we show that using \lang, we can achieve parity with C for linear
        algebra routines, whilst having much better static guarantees about the
        linearity and aliasing behaviour of our programs.
\end{itemize}
